---
title: "TidyTuesday_2025-05-13"
author: "Ishita Khanna"
date: "`r Sys.Date()`"
output: html_document
runtime: shiny
---

```{r}
library(tidyverse)
library(dplyr)

vesuvius <- read_csv("data/vesuvius.csv")
```

```{r exploring, include = FALSE}
glimpse(vesuvius)
```

<!-- making an input slider filter as a ui shiny element, range of years -->

```{r}
sliderInput("yearRange", "Select year:",
            min = 2011, max = 2025, value = c(2013, 2025), sep = "")
```

<!-- making an input select filter as a ui shiny element, their review status -->

```{r}
selectInput("review", "Review Level:",
            choices = c("All", "revised", "preliminary"), selected = "All")
```

# Seismic Activity Timeline

### Question:
*How has the number and severity of seismic events changed over the last decade?*

### Approach:
To explore this, I have chosen *Area and Line chart* where users can filter the time period and review level:
 - The **tan area** represents the number of seismic events per year.
 - The **red line** tracks the average magnitude each year.
 
### Answer we are looking for:
 - How has the frequency of seismic activity evolved over time?
 - Are higher magnitude events gradually becoming more or less frequent?