---
title: "TidyTuesday(06-04-25)"
author: "Ishita Khanna"
date: "`r Sys.Date()`"
output: html_document
runtime: shiny
---

```{r load_dataset, include=FALSE}
library(tidyverse)
library(readr)
library(lubridate)

nsf_data <- read_csv("data/nsf_terminations.csv")
```

```{r prepare_rows, include=FALSE}
nsf_data <- nsf_data %>%
  mutate(start = nsf_startdate,
         term = termination_letter_date,
         end = nsf_expected_end_date) %>%
  select(-nsf_startdate, -termination_letter_date, -nsf_expected_end_date)
```


```{r glimpse}
glimpse(nsf_data)

summary(nsf_data)

head(nsf_data)
```

# Grant termination Timeline

#### How long is the duration of NSF project are before they are terminated and how far were they from their expected end date 

To explore this I have chosen a *Gantt chart* as an interactive chart where user can use the directorate they want to see information about:
- starting point should be the project start date
- end point should be the termination letter received date 
- dotted line from the end date should be to the point which tells the end date of the project

#### Answer we are looking for:
- How early grants were terminated.
- Whether certain NSF directorates see earlier terminations than others.

```{r ui, echo=TRUE}
selectInput("dir", "Select a Directorate:",
            choices = unique(nsf_data$directorate),
            selected = unique(nsf_data$directorate)[1])
```



