---
title: "TidyTuesday(06-04-25)"
author: "Ishita Khanna"
date: "`r Sys.Date()`"
output: html_document
runtime: shiny
---

```{r load_dataset, include=FALSE}
library(tidyverse)
library(readr)
library(lubridate)
library(plotly)

nsf_data <- read_csv("data/nsf_terminations.csv")
```

```{r prepare_rows, include=FALSE}
nsf_data <- nsf_data %>%
  mutate(start = nsf_startdate,
         term = termination_letter_date,
         end = nsf_expected_end_date) %>%
  select(-nsf_startdate, -termination_letter_date, -nsf_expected_end_date)
```


```{r glimpse, include=FALSE}
glimpse(nsf_data)

summary(nsf_data)

head(nsf_data)
```

# Grant Termination Timeline

#### How long is the duration of NSF project are before they are terminated and how far were they from their expected end date 

To explore this I have chosen a *Gantt chart* as an interactive chart where user can use the directorate they want to see information about:
- starting point should be the project start date
- end point should be the termination letter received date 
- dotted line from the end date should be to the point which tells the end date of the project

#### Answer we are looking for:
- How early grants were terminated.
- Whether certain NSF directorates see earlier terminations than others.

```{r ui, echo=FALSE}
selectInput("dir", "Select a Directorate:",
            choices = unique(nsf_data$directorate),
            selected = unique(nsf_data$directorate)[1])
```

```{r gantt_plot, echo=FALSE}
# renderPlot({
#   filtered <- nsf_data %>%
#     filter(directorate == input$dir) %>%
#     arrange(start) %>%
#     slice_head(n = 10) %>%
#     mutate(project_id = row_number())
#   
# ggplot(filtered, aes(y = reorder(project_id, -project_id))) +
#   geom_segment(aes(x = start, xend = term, yend = project_id), color = "black") +
#   geom_segment(aes(x = term, xend = end, yend = project_id), linetype = "dotted", color = "gray50") +
#   geom_point(aes(x = start), color = "blue", size = 2) +
#   geom_point(aes(x = term), color = "red", size = 2) +
#   geom_point(aes(x = end), color = "darkgreen", size = 2) +
#   labs(title = paste("NSF Projects Timeline for", input$dir, "Directorate"),
#        x = "Date", y = "Projects") +
#   theme_minimal()
# })
```

```{r plotting, echo=FALSE}
renderPlotly({
  filtered <- nsf_data %>%
  filter(directorate == input$dir) %>%
  arrange(start) %>%
  slice_head(n = 30) %>%
  mutate(label = paste(org_name, " - ", substr(grant_number, 1, 5)))

p <- ggplot(filtered, aes(y = reorder(label, -start))) +
  geom_segment(aes(x = start, xend = term, y = label, yend = label, text = paste("Start:", start, "<br>Term:", term)), color = "black", size = 1) +
  geom_segment(aes(x = term, xend = end, y = label, yend = label, text = paste("Term:", term, "<br>End:", end)), linetype = "dotted", color = "gray50") +
  geom_point(aes(x = start, y = label, text = paste("Start:", start)), color = "blue", size = 2) +
  geom_point(aes(x = term, y = label, text = paste("Term:", term)), color = "red", size = 2) +
  geom_point(aes(x = end, y = label, text = paste("End:", end)), color = "darkgreen", size = 2) +
  labs(title = paste("Project Duration vs. Termination –", input$dir),
       x = "Date", y = "Projects") +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 6))

ggplotly(p, tooltip = "text")
})
```

