---
title: "Global Measles Outbreaks"
author: "Ishita Khanna"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: flatly
    highlight: tango
    df_print: paged
    fig_caption: true
---

<style>
/* Expand content width */
.main-container {
  max-width: 100% !important;
  padding: 0 30px;
}

/* Optional: improve figure size */
.plot-container {
  width: 100% !important;
}

.plot {
  margin: auto;
}

/* Optional: adjust figure captions */
.figure > figcaption {
  text-align: center;
  font-style: italic;
  color: #555;
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 12,
  fig.height = 8
)
```

```{r loading, include=FALSE}
library(tidyverse)
library(dplyr)
library(tidyr)
library(scales)
library(ggplot2)
library(plotly)
```

```{r dataset, include=FALSE}
cases_year <- read_csv("data/cases_year.csv")
cases_month <- read_csv("data/cases_month.csv")

glimpse(cases_year)
glimpse(cases_month)
```

```{=html}
<style>
body {
  font-family: "Roboto", sans-serif;
  line-height: 1.6;
  font-size: 16px;
}
</style>
```
::: {style="background-color:#e7f3fe; padding:15px; border-left:5px solid #2196F3;"}
<h4><strong>Project Overview</strong></h4>

<p>This analysis explores global measles trends using WHO surveillance data. It identifies high-risk countries, outbreak patterns, and under-reporting regions to aid data-driven public health decisions.</p>
:::

<br> The goal of this analysis is to:

-   **Analyse** monthly and seasonal outbreak patterns across countries.
-   **Visualise** how measles spread over time and geography.
-   **Evaluate** reporting and surveillance quality.
-   **Highlight** the preventable human impact of measles outbreaks.

```{r eda-summary, include=FALSE}
n_countries <- n_distinct(cases_year$country)
n_regions <- n_distinct(cases_year$region)

measles_median <- comma(median(cases_year$measles_total, na.rm = TRUE))
measles_max <- comma(max(cases_year$measles_total, na.rm = TRUE))

rubella_median <- comma(median(cases_year$rubella_total, na.rm = TRUE))
rubella_max <- comma(max(cases_year$rubella_total, na.rm = TRUE))
```

------------------------------------------------------------------------

## Summary of the Dataset

This dataset covers **`r n_countries` countries** across **`r n_regions` global regions**.

**Measles** case counts vary widely, with a **median of `r measles_median` cases** and a **maximum of `r measles_max`** reported in a single country-year.

**Rubella** case counts are much lower, with a **median of `r rubella_median` cases** (meaning half of country reported no cases at all) and a **maximum of `r rubella_max`**.

This highlights both the variability of outbreaks and under-reporting in some regions.

------------------------------------------------------------------------

## Monthly Outbreak Fingerprints (Top Countries)

To understand **seasonal patterns** in measles outbreaks, we visualise the monthly distribution of cases across the **12 most affected countries**.

This reveals outbreak "fingerprints" of how measles tends to spike during specific months in different regions.

```{r heatmap, echo = FALSE, fig.cap = "Figure 1: Monthly Measles Outbreak Patterns in the Top 12 Countries"}
top_countries <- cases_month %>%
  group_by(country) %>%
  summarise(total_cases = sum(measles_total, na.rm = TRUE)) %>%
  arrange(desc(total_cases)) %>%
  slice_head(n = 12) %>%
  pull(country)

heatmap_data <- cases_month %>%
  filter(country %in% top_countries) %>%
  group_by(country, month) %>%
  summarise(total_cases = sum(measles_total, na.rm = TRUE), .groups = "drop")

ggplot(heatmap_data, aes(x = factor(month), y = fct_reorder(country, -total_cases), fill = total_cases)) +
  geom_tile(color = "white", linewidth = 0.4) +
  scale_fill_gradient(low = "#fff5f0", high = "#cb181d", name = "cases") +
  labs(
    title = "Monthly Measles Outbreak Patterns",
    x = "Month",
    y = "Country"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "right"
  )
```

------------------------------------------------------------------------

## Bubble Chart: Coverage vs. Incidence

To better understand how **estimated vaccine coverage** and **measles incidence** relate across countries, this bubble chart compares the two metrics.

-   **X-axis**: Estimated vaccine coverage (95 - incidence/100)
-   **Y-axis**: Mean annual incidence (per million people)
-   **Bubble size**: Total population
-   **Color**: WHO region

This chart helps reveal **high-risk countries** those with **large populations**, **low vaccine coverage**, and **high incidence**, which should be prioritised for vaccination and surveillance efforts.

```{r bubble_data, include=FALSE, warning=FALSE}
bubble_df <- cases_year %>%
  group_by(country, region) %>%
  summarise(
    mean_incidence = mean(measles_incidence_rate_per_1000000_total_population, na.rm = TRUE),
    total_population = mean(total_population, na.rm = TRUE)
  ) %>%
  filter(!is.na(mean_incidence), mean_incidence > 0, total_population > 0) %>%
  mutate(
    vaccine_coverage = 95 - (mean_incidence / 100), # derived estimate
    bubble_size = sqrt(total_population) / 1000     # scaled for plotting
  )

bubble_df <- bubble_df %>%
  mutate(
    hover_text = paste0(
      "Country: ", country, "\n",
      "Region: ", region, "\n",
      "Incidence: ", round(mean_incidence, 1), "\n",
      "Coverage: ", round(vaccine_coverage, 1), "%\n",
      "Population: ", scales::comma(round(total_population))
    )
  )
```

```{r bubbe_chart, echo=FALSE, warning=FALSE, results='asis'}
bubble_text <- ggplot(bubble_df, aes(x = vaccine_coverage, y = mean_incidence)) +
  geom_point(aes(size = bubble_size, color = region, text = hover_text), alpha = 0.7)  +
  scale_size_continuous(range = c(2, 12), name = "Population\n(scaled approx)") +
  scale_color_brewer(palette = "Set2") +
  labs(
    title = "Measles Incidence vs. Estimated Vaccine Coverage",
    subtitle = "Bubble size represents population (Top countries by incidence)",
    x = "Estimated Vaccine Coverage (%)",
    y = "Mean Measles Incidence per Million",
    color = "Region"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11)
  )

htmltools::tags$div(
  style = "display: flex; justify-content: center;",
  ggplotly(bubble_text, tooltip = "text")
)

cat("<div style='text-align:center; font-style:italic; margin-top: 8px;'>
Figure 2: Measles Incidence vs. Estimated Vaccine Coverage (Bubble size = Population)
</div>")
```

------------------------------------------------------------------------

## Scatter Plot: Reported Cases vs Incidence Rate

This plot compares **total reported measles cases** with **average incidence rate per million**, colored by WHO region.

-   **Top-left**: High incidence but low total cases, possible *under-reporting*.
-   **Top-right**: High incidence and high totals, likely *accurate reporting*.

This helps identify **gaps in disease surveillance** across countries.

```{r scatter, echo=FALSE, warning=FALSE, fig.cap="Figure 3: Top-left = high incidence, low reporting | Top-right = high incidence & reporting"}
cases_year %>%
  group_by(country, region) %>%
  summarise(
    total_cases = sum(measles_total, na.rm = TRUE),
    avg_incidence = mean(measles_incidence_rate_per_1000000_total_population, na.rm = TRUE),
    avg_population = mean(total_population, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(avg_incidence > 0, total_cases > 0) %>%
  ggplot(aes(x = total_cases, y = avg_incidence, label = country, color = region)) +
  geom_point(alpha = 0.7) +
  geom_text(check_overlap = TRUE, size = 3, nudge_x = 0.1) +
  scale_x_log10(labels = comma) +
  scale_y_continuous(labels = comma) +
  scale_color_brewer(palette = "Set2") +
  labs(
    title = "Mismatch Between Reported Cases and Incidence Rate",
    subtitle = "Log-scale reveals anomalies: High incidence but low total cases may indicate under-reporting",
    x = "Total Measles Cases (log scale)",
    y = "Average Incidence per Million",
    color = "WHO Region"
  ) +
  coord_cartesian(clip = "off") +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11),
    legend.position = "right",
    plot.margin = margin(12,20,10,20)
  )
```

------------------------------------------------------------------------
